---
title: "Interpretation and Analysis of the FIP Matrix"
subtitle: "PARC-HBM and PARCToxicology as FAIR Implementing Communities"
author: "Viviane Girardin"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 6
)

# --- Packages -------------------------------------------------------------
# Install any missing packages, then load them.
required <- c("tidyverse", "knitr", "scales")
to_install <- required[!required %in% installed.packages()[, "Package"]]
if (length(to_install) > 0) {
  install.packages(to_install, repos = "https://cloud.r-project.org")
}

library(tidyverse)
library(knitr)
library(scales)

# kableExtra gives prettier tables but is optional here. If it is installed we
# use it; otherwise we fall back to knitr::kable via thin shims so the rest of
# the document needs no changes.
if (requireNamespace("kableExtra", quietly = TRUE)) {
  library(kableExtra)
} else {
  kbl           <- function(x, ...) knitr::kable(x, ...)
  kable_styling <- function(x, ...) x
  scroll_box    <- function(x, ...) x
}

# Render kable tables as HTML so they can be embedded in collapsible blocks.
options(knitr.table.format = "html")

# Wrap a (k)able in a collapsible <details> block. Works in static HTML, so it
# survives publishing to GitHub Pages with no JavaScript required.
collapsible <- function(tbl, summary = "Show table", open = FALSE) {
  knitr::asis_output(paste0(
    if (open) "<details open>" else "<details>",
    "\n<summary>", summary, "</summary>\n\n",
    paste(as.character(tbl), collapse = "\n"),
    "\n\n</details>"
  ))
}
```

# Background

Through a **FAIR Implementation Profile (FIP)**, a **FAIR Implementing Community
(FIC)** declares the **FAIR Enabling Resources (FERs)** (standards, identifiers,
protocols, vocabularies, licences, etc) that it uses to address the FAIR
principles. The FIP is built around **21 questions** derived from the **14 FAIR
principles** (Findable, Accessible, Interoperable, Reusable). A **FIP matrix** places two or more FICs side by side so their FER choices can be compared.

This report interprets a FIP matrix for two FICs within the **PARC** project:

- **PARC-HBM** — Human Biomonitoring
- **PARCToxicology**

These were the only FICs related to PARC found in https://fip-matrix.streamlit.app/.

For each FER, both FICs are scored on a **0–3 scale**:

```{r legend-table, echo=FALSE}
legend_df <- tibble(
  Score = c(0L, 1L, 2L, 3L),
  Meaning = c(
    "No data",
    "Resource in development / future use",
    "Available resource / future use",
    "Available resource / current use"
  )
)

legend_df %>%
  kbl(caption = "Scoring legend used throughout this report") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("Scoring legend (0–3 scale)", open = TRUE)
```

<!-- # Data Import and Preparation -->

```{r load-data}
# Adjust the path if the file lives elsewhere.
csv_path <- "2026-06-23T07-20_export.csv"

raw <- read_csv(csv_path, show_col_types = FALSE)

# Tidy column names for convenience.
fip <- raw %>%
  rename(
    FIP_question = `FIP questions`,
    Resource     = `FAIR Supporting Resource`,
    PARC_HBM     = `PARC-HBM`,
    PARC_Tox     = `PARCToxicology`
  ) %>%
  mutate(
    Resource = str_squish(Resource),
    # Derive the top-level FAIR principle from the first letter of the question.
    Principle = recode(str_sub(FIP_question, 1, 1),
      F = "Findable",
      A = "Accessible",
      I = "Interoperable",
      R = "Reusable"
    ),
    # Derive whether the question concerns Data (-D) or Metadata (-MD).
    Target = case_when(
      str_detect(FIP_question, "-MD$") ~ "Metadata",
      str_detect(FIP_question, "-D$")  ~ "Data",
      TRUE                             ~ "General"
    )
  )
```

```{r score-labels}
# Ordered factor for the maturity scale, reused in plots and tables.
score_levels  <- 0:3
score_labels  <- c("No data",
                    "In development",
                    "Available / future",
                    "Available / current")

label_score <- function(x) {
  factor(x, levels = score_levels, labels = score_labels)
}
```

# Data Overview

```{r overview-counts}
overview <- tibble(
  Metric = c("Total matrix rows (FER × question)",
             "Distinct FERs",
             "Distinct FIP questions",
             "FAIR principles covered"),
  Value = c(nrow(fip),
            n_distinct(fip$Resource),
            n_distinct(fip$FIP_question),
            n_distinct(fip$Principle))
)

overview %>%
  kbl(caption = "High-level overview of the FIP matrix") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("High-level overview")
```

The matrix is organised by FAIR principle. The table below shows how many rows
fall under each principle and whether they target **Data** or **Metadata**.

```{r principle-target-breakdown}
fip %>%
  count(Principle, Target) %>%
  pivot_wider(names_from = Target, values_from = n, values_fill = 0) %>%
  arrange(factor(Principle,
                 levels = c("Findable", "Accessible",
                            "Interoperable", "Reusable"))) %>%
  kbl(caption = "Number of matrix rows by FAIR principle and target") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("Rows by FAIR principle and target")
```

# Summary Statistics per FIC

The matrix is reshaped into long form so the two FICs can be compared directly.

```{r long-form}
fip_long <- fip %>%
  pivot_longer(
    cols = c(PARC_HBM, PARC_Tox),
    names_to = "Community",
    values_to = "Score"
  ) %>%
  mutate(
    Community = recode(Community,
      PARC_HBM = "PARC-HBM",
      PARC_Tox = "PARCToxicology"),
    Score_label = label_score(Score)
  )
```

```{r score-distribution-table}
fip_long %>%
  count(Community, Score_label) %>%
  pivot_wider(names_from = Score_label, values_from = n, values_fill = 0) %>%
  kbl(caption = "Distribution of scores per FIC") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("Score distribution per FIC")
```

```{r score-distribution-plot, fig.cap="Distribution of maturity scores by community."}
ggplot(fip_long, aes(x = Score_label, fill = Community)) +
  geom_bar(position = position_dodge(preserve = "single")) +
  scale_fill_manual(values = c("PARC-HBM" = "#2c7fb8",
                               "PARCToxicology" = "#de2d26")) +
  labs(x = "Maturity / adoption level", y = "Number of resources",
       title = "Resource maturity distribution",
       fill = "Community") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r adoption-summary}
# An "adopted" resource is one that is at least available (score >= 2).
adoption <- fip_long %>%
  group_by(Community) %>%
  summarise(
    `Mean score`              = round(mean(Score), 2),
    `No data (0)`             = sum(Score == 0),
    `In development (1)`      = sum(Score == 1),
    `Available, future (2)`   = sum(Score == 2),
    `Available, current (3)`  = sum(Score == 3),
    `Currently in use (3) %`  = percent(mean(Score == 3), accuracy = 0.1),
    .groups = "drop"
  )

adoption %>%
  kbl(caption = "Score summary per FIC") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("Score summary per FIC")
```

# The FIP Matrix as a Heatmap

In the heatmap below each cell is a FER, with colour showing its score for each
FIC. FERs are grouped by FAIR principle.

```{r heatmap, fig.height=14, fig.width=10, fig.cap="FIP matrix heatmap. Rows are resources (grouped by FAIR principle); columns are communities; colour encodes maturity."}
heat_df <- fip_long %>%
  mutate(
    # Build a readable row label combining question and resource.
    row_label = paste0(FIP_question, " | ", str_trunc(Resource, 45))
  )

ggplot(heat_df, aes(x = Community, y = fct_rev(factor(row_label)), fill = Score_label)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = Score), size = 3, colour = "grey20") +
  scale_fill_manual(
    values = c(
      "No data"              = "#f0f0f0",
      "In development"       = "#fee08b",
      "Available / future"   = "#91cf60",
      "Available / current"  = "#1a9850"
    ),
    drop = FALSE,
    name = "Maturity"
  ) +
  facet_grid(Principle ~ ., scales = "free_y", space = "free_y", switch = "y") +
  labs(x = NULL, y = NULL, title = "FIP Matrix Heatmap") +
  theme_minimal(base_size = 9) +
  theme(
    strip.text.y.left = element_text(angle = 0, face = "bold"),
    strip.placement = "outside",
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    legend.position = "top"
  )
```

# Comparison Between the Two FICs

For every FER, the two FICs are compared and the relationship classified as
shared, specific to one FIC, emerging, or absent from both.

```{r gap-analysis}
comparison <- fip %>%
  mutate(
    status = case_when(
      PARC_HBM == 0 & PARC_Tox == 0          ~ "Neither (no data)",
      PARC_HBM >= 2 & PARC_Tox >= 2          ~ "Shared / aligned",
      PARC_HBM >= 2 & PARC_Tox <  2          ~ "HBM-only strength",
      PARC_Tox >= 2 & PARC_HBM <  2          ~ "Tox-only strength",
      TRUE                                   ~ "Emerging (in development)"
    )
  )

comparison %>%
  count(status, name = "Resources") %>%
  arrange(desc(Resources)) %>%
  kbl(caption = "Relationship between the two FICs, per FER") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("Relationship counts per FER")
```

```{r comparison-plot, fig.cap="Number of resources by community-relationship category."}
comparison %>%
  count(status) %>%
  ggplot(aes(x = reorder(status, n), y = n, fill = status)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  expand_limits(y = max(table(comparison$status)) * 1.1) +
  labs(x = NULL, y = "Number of resources",
       title = "Alignment and divergence between PARC-HBM and PARCToxicology") +
  theme_minimal(base_size = 12)
```

## FERs shared by both FICs

```{r shared-resources}
comparison %>%
  filter(status == "Shared / aligned") %>%
  select(FIP_question, Principle, Resource, PARC_HBM, PARC_Tox) %>%
  arrange(Principle, FIP_question) %>%
  kbl(caption = "FERs available to both FICs (score >= 2 each)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("FERs shared by both FICs")
```

# Coverage by FAIR Principle

The mean score per principle shows which FAIR dimensions each FIC supports best.

```{r principle-coverage}
principle_summary <- fip_long %>%
  group_by(Principle, Community) %>%
  summarise(mean_score = mean(Score), .groups = "drop")

ggplot(principle_summary,
       aes(x = factor(Principle,
                      levels = c("Findable", "Accessible",
                                 "Interoperable", "Reusable")),
           y = mean_score, fill = Community)) +
  geom_col(position = position_dodge()) +
  geom_text(aes(label = round(mean_score, 1)),
            position = position_dodge(width = 0.9), vjust = -0.3, size = 3.2) +
  scale_fill_manual(values = c("PARC-HBM" = "#2c7fb8",
                               "PARCToxicology" = "#de2d26")) +
  ylim(0, 3) +
  labs(x = "FAIR principle", y = "Mean maturity score (0–3)",
       title = "FAIR coverage by principle and community",
       fill = "Community") +
  theme_minimal(base_size = 12)
```

```{r principle-coverage-table}
principle_summary %>%
  pivot_wider(names_from = Community, values_from = mean_score) %>%
  mutate(across(where(is.numeric), ~round(.x, 2))) %>%
  arrange(factor(Principle,
                 levels = c("Findable", "Accessible",
                            "Interoperable", "Reusable"))) %>%
  kbl(caption = "Mean maturity score per FAIR principle") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("Mean score per FAIR principle")
```

# Data vs. Metadata

Many FIP questions are declared separately for **Data (-D)** and **Metadata
(-MD)**. Comparing the two shows whether each FIC supports its data and metadata
to a similar degree.

```{r data-vs-metadata}
fip_long %>%
  filter(Target %in% c("Data", "Metadata")) %>%
  group_by(Community, Target) %>%
  summarise(mean_score = round(mean(Score), 2), .groups = "drop") %>%
  pivot_wider(names_from = Target, values_from = mean_score) %>%
  kbl(caption = "Mean maturity for Data vs. Metadata targets") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("Data vs. Metadata mean scores")
```

# Interpretation and Key Findings

```{r key-numbers, include=FALSE}
# Pull a few numbers so the narrative below stays in sync with the data.
hbm_mean   <- round(mean(fip$PARC_HBM), 2)
tox_mean   <- round(mean(fip$PARC_Tox), 2)
n_shared   <- sum(comparison$status == "Shared / aligned")
n_hbm_only <- sum(comparison$status == "HBM-only strength")
n_tox_only <- sum(comparison$status == "Tox-only strength")
```

A few points stand out from the matrix (`r nrow(fip)` FER entries across
`r n_distinct(fip$Principle)` FAIR principles):

- Scores are modest and broadly similar between the two FICs (mean
  **`r hbm_mean`** for PARC-HBM and **`r tox_mean`** for PARCToxicology). Many
  entries are `0`, so both FIPs are still being built up rather than complete.

- Overlap is limited: only **`r n_shared`** FER(s) are available in both FICs,
  while **`r n_hbm_only`** are specific to PARC-HBM and **`r n_tox_only`** to
  PARCToxicology. The two FICs largely address the same questions with
  different FERs.

- The difference follows the domains. PARCToxicology uses chemistry-specific
  FERs (e.g. InChI, IUPAC Gold Book, ChEMBL), while PARC-HBM uses
  catalogue and infrastructure FERs (e.g. CSW, CKAN, Keycloak, CEDAR, RDF).
  The Interoperable and Reusable principles diverge most, since these rely on
  shared vocabularies, identifiers, and licences.

As long as the FERs are not fully machine-actionable and FAIR, these
differences in FER choice are worth documenting: they mark the points where
re-use of (meta)data between the two FICs would need a case-by-case check or an
agreed mapping.

# Appendix

## FERs specific to one FIC

FERs available in one FIC but not the other (score >= 2 in one, < 2 in the
other). Because a score of 0 means "no data", some of these gaps may reflect
missing declarations rather than a deliberate difference in FER use.

```{r divergent-resources}
comparison %>%
  filter(status %in% c("HBM-only strength", "Tox-only strength")) %>%
  select(FIP_question, Principle, Resource, PARC_HBM, PARC_Tox, status) %>%
  arrange(status, Principle, FIP_question) %>%
  kbl(caption = "FERs available in one FIC but not the other") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  collapsible("FERs specific to one FIC")
```

## Full matrix

```{r full-matrix}
fip %>%
  select(FIP_question, Principle, Target, Resource, PARC_HBM, PARC_Tox, Link) %>%
  arrange(Principle, FIP_question, Resource) %>%
  kbl(caption = "Complete FIP matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 11) %>%
  scroll_box(height = "500px") %>%
  collapsible("Complete FIP matrix (all 51 rows)")
```
